---
title: "Google Data Analytics Capstone Project - Bellabeat"
author: "Lin Jiaming"
date: "7 June 2021"
output: html_document
---

## Description

**Activity:** Data Analysis, Data Visualisation  
**Dataset:** FitBit Fitness Tracker Data (Kaggle)  
**Table:** dailyActivity_clean.csv, sleepDay_clean.csv  

### Load Libraries

```{r library, message=FALSE}
library(tidyverse)
library(ggplot2)
library(lubridate)
```

### Import Data

```{r import data, message=FALSE}
activity_df = read_csv('dailyActivity_clean.csv')
sleep_df = read_csv('sleepDay_clean.csv')
```

### Inspect Data

Inspect column header and data type only  

Data is cleaned prior this activity, refer to:  

* Bellabeat_dailyActivity_clean.html
* Bellabeat_sleepDay_clean.html

```{r inspect column header + data type}
glimpse(activity_df)
glimpse(sleep_df)
```
```{r data type update}
# Update data types to factor, for visualisation purposes
activity_df$id <- factor(activity_df$id)
sleep_df$id <- factor(sleep_df$id)
sleep_df$total_sleep_records <- factor(sleep_df$total_sleep_records)
```

## Summary Descriptions

**Activity Table:**

* Total records = 940
* Unique users = 33
* Unique dates = 31
* Start date = 2016-04-12  
* End date = 2016-05-12

**Sleep Table:**

* Total records = 410
* Unique users = 24
* Unique dates = 31
* Start date = 2016-04-12  
* End date = 2016-05-12

```{r summary}
activity_df %>% summarise(record_cnt = nrow(.), 
                          user_cnt = n_distinct(.$id), 
                          date_cnt = n_distinct(.$activity_date), 
                          user_date_cnt = n_distinct(.$id, .$activity_date),
                          start_date = min(.$activity_date), 
                          end_date = max(.$activity_date))

sleep_df %>% summarise(record_cnt = nrow(.), 
                       user_cnt = n_distinct(.$id), 
                       date_cnt = n_distinct(.$sleep_day), 
                       user_date_cnt = n_distinct(.$id, .$sleep_day),
                       start_date = min(.$sleep_day), 
                       end_date = max(.$sleep_day))
```

## Visualisations

#### User Activity: Tracking *Total Steps*

33 users tracked their data  
21 out of 33 users (64%) tracked their data for **all days** -- *see non-blank rows in visual*  
All users tracked their data in **running days**

```{r user activity heat map 1, echo=FALSE}
# Y-axis shows the id of all 33 users
ggplot(activity_df) + 
  geom_tile(aes(activity_date, id, fill = total_steps)) +
  scale_fill_continuous(trans = 'reverse') +                # reverse fill scale
  theme(panel.grid = element_blank(),
        panel.background = element_blank()) + 
  labs(title = 'User Steps per Day', 
       subtitle = "Every user's steps for each day",
       x = 'Date', y = 'User',
       fill = 'Steps') +                                    # rename legend
  guides(fill = guide_legend(reverse = TRUE)) +             # reverse legend
  scale_x_date(date_breaks = '4 day', 
               date_labels = '%d-%b',
               expand = c(0,0)) +                           # fill gap between axis and plot
  scale_y_discrete(limits = rev)                            # reverse y-axis
```
```{r}
activity_df %>% 
  group_by(id) %>% 
  summarise(days_tracked = length(activity_date)) %>% 
  group_by(days_tracked) %>% 
  summarise(user_count = length(id)) %>% 
  arrange(-days_tracked)
```
--------------------------------------------------------------------------------

#### User Activity: Tracking *Total Steps*

77 out of 940 records (8%) registered **total steps of 0** for the entire day

```{r user activity heat map 2, echo=FALSE}
# Convert steps to discrete variable
activity_df_steps <- activity_df %>%
  mutate(category = if_else(total_steps == 0, '= 0', '> 0'))

# Y-axis shows the id of all 33 users
ggplot(activity_df_steps) + 
  geom_tile(aes(activity_date, id, fill = category)) +
  scale_fill_manual(values = c('#e15759', '#4e79a7')) +
  theme(panel.grid = element_blank(),
        panel.background = element_blank()) + 
  labs(title = 'User Steps per Day', 
       subtitle = "Every user's steps for each day",
       x = 'Date', y = 'User',
       fill = 'Steps') +                                # rename legend
  scale_x_date(date_breaks = '4 day', 
               date_labels = '%d-%b',
               expand = c(0,0)) +                       # fill gap between axis and plot
  scale_y_discrete(limits = rev)                        # reverse y-axis
```
```{r}
activity_df %>% 
  filter(total_steps == 0) %>% 
  nrow()
```

--------------------------------------------------------------------------------

#### User Activity: Tracking *Sleep*

24 out of 33 users (73%) tracked their sleep  
3 out of 33 users (9%) tracked their sleep data for **all days** -- *see non-blank rows in visual*  
5 out of 33 users (15%) tracked their sleep data in **running days**

```{r user activity heat map 3, echo=FALSE}
# Y-axis shows the id of 24 users who tracked their sleep
ggplot(sleep_df) + 
  geom_tile(aes(sleep_day, id, fill = total_minutes_asleep/60)) +
  scale_fill_continuous(trans = 'reverse') +            # reverse fill scale
  theme(panel.grid = element_blank(),
        panel.background = element_blank()) + 
  labs(title = 'User Sleep Hours per Day', 
       subtitle = "Every user's sleep hours for each day",
       x = 'Date', y = 'User',
       fill = 'Sleep Hours') +                          # rename legend
  guides(fill = guide_legend(reverse = TRUE)) +         # reverse legend
  scale_x_date(date_breaks = '4 day', 
               date_labels = '%d-%b',
               expand = c(0,0)) +                       # fill gap between axis and plot
  scale_y_discrete(limits = rev)                        # reverse y-axis
```
```{r}
sleep_df %>% 
  group_by(id) %>% 
  summarise(days_tracked = length(sleep_day)) %>% 
  group_by(days_tracked) %>% 
  summarise(user_count = length(id)) %>% 
  arrange(-days_tracked)
```
```{r}
# 24 users tracked their sleep data
# 5 users tracked their sleep data in running days
# 19 users did not track their sleep data in running days
sleep_df %>% 
  group_by(id) %>% 
  summarise(date_diff = sleep_day - lag(sleep_day, 1)) %>% 
  filter(date_diff != 1) %>% 
  distinct(id) %>% 
  nrow()
```

--------------------------------------------------------------------------------

#### Filter Data

The **following visuals** will use **filtered data**, whereby records with total steps = 0 or total distance = 0 are removed.

```{r filter data}
activity_filter_df <- activity_df %>% 
  filter(total_steps != 0 | total_distance != 0)
```

--------------------------------------------------------------------------------

#### User Profile: Users by *Steps*

* 7 inactive users (21%)
* 19 active users (58%)
* 7 very active users (21%)

```{r user profile pie chart, echo=FALSE}
activity_df_profile <- activity_filter_df %>% 
  group_by(id) %>% 
  summarise(user_avg_steps = mean(total_steps)) %>%                         # average steps for each user
  mutate(user_profile = case_when(user_avg_steps < 5000 ~ 'Inactive',
                                  user_avg_steps >= 10000 ~ 'Very Active',
                                  TRUE ~ 'Active')) %>%                     # classify each user profile
  group_by(user_profile) %>% 
  summarise(user_count = n_distinct(id)) %>% 
  mutate(percent = round(user_count / sum(.$user_count) * 100)) %>%         # calculate percentage
  mutate(y_position = 100 - (cumsum(percent) - 0.5*percent) + 2)            # position of each pie segment 

ggplot(activity_df_profile) + 
  geom_bar(aes('', percent, fill = user_profile), 
           stat = 'identity',
           width = 1,  
           colour = 'white') + 
  coord_polar('y', start = 0) +                                             # create pie chart
  scale_fill_manual(values = c('#4e79a7', '#e15759', '#59a14f')) +
  theme_void() +
  theme(legend.position = 'none',
        plot.caption = element_text(hjust = 0)) +
  labs(title = 'User Profile by Steps', 
       subtitle = 'Steps are averaged individually for each user',
       caption = 'Inactive: less than 5,000 steps per day \nActive: between 5,000 and 9,999 steps per day \nVery Active: at least 10,000 steps per day') +
  geom_text(aes('', y = y_position, label = paste0(user_profile,'\n',percent,'%')), 
            color = 'white', 
            size = 5)
```

--------------------------------------------------------------------------------

#### User Profile: Users by *Steps*

There is a **wide and normal distribution** of inactive, active, and very active users  
The average user takes an average of **7928 steps daily**, below the recommended 10,000 steps daily ([Reference](https://www.healthline.com/health/how-many-steps-a-day#Why-10,000-steps?
))

```{r user profile bar chart 1, echo=FALSE}
activity_df_user_steps <- activity_filter_df %>% 
  group_by(id) %>% 
  summarise(user_avg_steps = mean(total_steps)) %>%                 # average steps for each user
  mutate(user_profile = case_when(user_avg_steps < 5000 ~ 'Inactive',
                                  user_avg_steps >= 10000 ~ 'Very Active',
                                  TRUE ~ 'Active'))                 # classify each user

ggplot(activity_df_user_steps) + 
  geom_bar(aes(reorder(id, user_avg_steps), user_avg_steps,         # reorder bars by steps
               fill = factor(user_profile, levels = c('Inactive', 'Active', 'Very Active'))), 
           stat = 'identity', 
           width = 1, 
           colour = 'white') +                                      # reorder legend by refactor user_profile
  scale_fill_manual(values = c('#e15759', '#4e79a7', '#59a14f')) + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.title = element_blank(),
        plot.caption = element_text(hjust = 0)) + 
  labs(title = 'Average Daily Total Steps per User', 
       subtitle = 'Steps are averaged individually for each user',
       x = 'User', y = 'Average Daily Total Steps',
       caption = 'Inactive: less than 5,000 steps per day \nActive: between 5,000 and 9,999 steps per day \nVery Active: at least 10,000 steps per day') +
  scale_y_continuous(expand = c(0,0)) +                             # fill gap between axis and plot
  geom_hline(aes(yintercept = mean(user_avg_steps))) +              # add average line
  geom_text(aes(min(as.numeric(id)), mean(user_avg_steps), 
                label = paste0('Average = ', round(mean(user_avg_steps)))), 
            hjust = 0, vjust = -0.4) +
  geom_hline(aes(yintercept = 10000)) +                             # add constant line
  geom_text(aes(min(as.numeric(id)), 10000, 
                label = paste0('Target = ', 10000)), 
            hjust = 0, vjust = -0.4)
# To-do: Add reference band using geom_rect()
```

--------------------------------------------------------------------------------

#### User Profile: Users by *Exercise Minutes*

There is a **wide distribution** of users  
The average user exercises an average of **34 minutes weekly**, below the recommended 150 minutes daily ([Reference](https://www.cdc.gov/physicalactivity/basics/adults))

```{r user profile bar chart 2, echo=FALSE, message=FALSE}
activity_df_user_mins <- activity_filter_df %>% 
  mutate(ex_mins = very_active_minutes + fairly_active_minutes,
         week = epiweek(.$activity_date)) %>%                       # categorise by day of week
  group_by(id, week) %>% 
  summarise(ex_mins_weekly = mean(ex_mins)) %>%                     # average weekly exercise minutes
  group_by(id) %>% 
  summarise(ex_mins_user = mean(ex_mins_weekly))                    # average weekly exercise minutes per user

ggplot(activity_df_user_mins) + 
  geom_bar(aes(id, ex_mins_user), 
           stat = 'identity', 
           width = 1, 
           colour = 'white',
           fill = '#4e79a7') +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        plot.caption = element_text(hjust = 0))+ 
  labs(title = 'Average Weekly Exercise Minutes per User', 
       subtitle = 'Exercise minutes are averaged individually for each user',
       caption = 'Exercise Minutes = Very Active Minutes + Fairly Active Minutes',
       x = 'User', y = 'Average Weekly Exercise Minutes') +
  scale_y_continuous(expand = c(0,0), limits = c(0,160)) +          # increase axis limits
  geom_hline(aes(yintercept = mean(ex_mins_user))) +                # add average line
  geom_text(aes(min(as.numeric(id)), mean(ex_mins_user), 
                label = paste0('Average = ', round(mean(ex_mins_user)))), 
            hjust = 0, vjust = -0.4) +
  geom_hline(aes(yintercept = 150)) +                               # add constant line
  geom_text(aes(min(as.numeric(id)), 150, 
                label = paste0('Target = ', 150)), 
            hjust = 0, vjust = -0.4)
```

--------------------------------------------------------------------------------

#### User Profile: Users by *Time to Fall Asleep*

Majority of the users (91%) have **no sleep issues**, and take an average of 25 minutes to fall asleep

```{r user profile bar chart 3, echo=FALSE}
sleep_df_user <- sleep_df %>% 
  mutate(fall_asleep_time = total_time_in_bed - total_minutes_asleep) %>% 
  group_by(id) %>% 
  summarise(fall_asleep_time_user = mean(fall_asleep_time)) %>% 
  mutate(category = if_else(fall_asleep_time_user > 90, 'Have Sleep Issues', 'No Sleep Issues'))

ggplot(sleep_df_user) + 
  geom_bar(aes(reorder(id, fall_asleep_time_user), fall_asleep_time_user, fill = category), 
           stat = 'identity', 
           width = 1, 
           colour = 'white') +
  scale_fill_manual(values = c('#e15759', '#4e79a7')) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), 
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.title = element_blank(),
        plot.caption = element_text(hjust = 0)) + 
  labs(title = 'Time to Fall Asleep per User', 
       subtitle = 'Time to fall asleep is averaged individually for each user',
       caption = 'Users with sleep issues, require more than 90 minutes to fall asleep \nTime to Fall Asleep in Bed = Total Time In Bed - Total Minutes Asleep',
       x = 'User', y = 'Time to Fall Asleep') +
  scale_y_continuous(expand = c(0,0)) +                                   # fill gap between axis and plot
  geom_hline(aes(yintercept = mean(fall_asleep_time_user)),               # add average line
             subset(sleep_df_user, category == 'Have Sleep Issues')) +    # use subset of data
  geom_text(aes(0, mean(fall_asleep_time_user), 
                label = paste0('Average = ', round(mean(fall_asleep_time_user)))),
            subset(sleep_df_user, category == 'Have Sleep Issues'), 
            hjust = -0.1, vjust = -29) +
  geom_hline(aes(yintercept = mean(fall_asleep_time_user)), 
             subset(sleep_df_user, category == 'No Sleep Issues')) + 
  geom_text(aes(0, mean(fall_asleep_time_user), 
                label = paste0('Average = ', round(mean(fall_asleep_time_user)))),
            subset(sleep_df_user, category == 'No Sleep Issues'), 
            hjust = -0.1, vjust = -16) +
  coord_flip()
```

--------------------------------------------------------------------------------

## Appendix

#### User Habit: *Activity* by Day of Week

Users are **most active** on Tuesdays and Saturdays  
Users are **least active** on Sundays  
Trend for active minutes and total steps is generally **similar**, except for Friday  

```{r user profile bar chart 4, echo=FALSE}
activity_df_wd <- activity_filter_df %>% 
  mutate(day_of_week = wday(activity_date, label = TRUE, abbr = TRUE)) %>% 
  group_by(day_of_week) %>% 
  summarise(avg_steps = round(mean(total_steps)), 
            avg_mins = round(mean(very_active_minutes + fairly_active_minutes + lightly_active_minutes)))

ggplot(activity_df_wd) + 
  geom_bar(aes(reorder(day_of_week, -avg_steps), avg_steps, fill = 'Steps'),    # create legend by fill
           stat = 'identity') +
  geom_line(aes(reorder(day_of_week, -avg_steps), 20*avg_mins, colour = 'Active Minutes'), 
            stat = 'identity', 
            group = 1, 
            size = 1.2) +
  geom_point(aes(reorder(day_of_week, -avg_steps), 20*avg_mins), 
             colour = '#f28e2b', size = 2) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_text(size = 11),
        axis.text.y = element_blank(),
        plot.caption = element_text(hjust = 0),
        legend.key = element_blank(),
        legend.title = element_blank()) + 
  labs(title = 'Activity by Day of Week', 
       subtitle = 'Measure is averaged across all user records',
       caption = 'Active Minutes = Very Active Minutes + Fairly Active Minutes + Lightly Active Minutes') +
  scale_x_discrete(expand = c(0,0)) +                                           # fill gap btw axis and plot
  scale_y_continuous(expand = c(0,0), sec.axis = sec_axis(~./20)) +             # add secondary axis and scale
  scale_fill_manual(values = c('Steps' = '#4e79a7')) +
  scale_color_manual(values = c('Active Minutes' = '#f28e2b')) +                # create legend for combo chart
  geom_text(aes(day_of_week, avg_steps, label = avg_steps), 
            colour = 'white', size = 5, vjust = 2) + 
  geom_text(aes(day_of_week, avg_mins, label = avg_mins), 
            colour = 'black', size = 5, vjust = -12)
```

--------------------------------------------------------------------------------

#### User Habit: *Sleep Hours* by Day of Week

Users **sleep the most** on days transitioning to Sunday (~20-50 mins more)  
Users **sleep the least** on days transitioning to Tuesday and Thursday (6.7 hours)  

```{r user profile bar chart 5, echo=FALSE}
sleep_df_wd <- sleep_df %>% 
  mutate(day_of_week = wday(sleep_day, label = TRUE, abbr = TRUE)) %>% 
  group_by(day_of_week) %>% 
  summarise(avg_sleep_time = round(mean(total_minutes_asleep)/60,1))

ggplot(sleep_df_wd) + 
  geom_bar(aes(reorder(day_of_week, -avg_sleep_time), avg_sleep_time),       # reorder bars by DOW
           stat = 'identity',
           fill = '#4e79a7') +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_text(size = 11),
        axis.text.y = element_blank()) + 
  labs(title = 'Sleep Hours by Day of Week', 
       subtitle = 'Measure is averaged across all user records',
       y = 'Total Hours Asleep') +
  scale_x_discrete(expand = c(0,0)) +                                        # fill gap between axis and plot
  scale_y_continuous(expand = c(0,0)) + 
  geom_text(aes(day_of_week, avg_sleep_time, label = avg_sleep_time), 
            colour = 'white', size = 5, vjust = 2)
```

--------------------------------------------------------------------------------

#### Trend: Steps vs Distance

There is **near perfect correlation** that higher total steps leads to higher distance covered  
Correlation coefficient of **0.98**  
* The closer to 1.0, the stronger the positive correlation  

```{r user activity scatter plot 1, echo=FALSE}
ggplot(activity_filter_df) + 
  geom_point(aes(total_steps, total_distance)) +
  geom_smooth(aes(total_steps, total_distance), 
              method = 'loess', formula = y~x) + 
  labs(title = 'Steps vs Distance', 
       subtitle = 'Correlation between steps and distance',
       x = 'Total Steps', y = 'Total Distance (km)')
```
```{r}
cor(activity_filter_df$total_steps, activity_filter_df$total_distance) %>% 
  round(2)
```

--------------------------------------------------------------------------------

#### Trend: Steps vs Calories

There is a **moderate correlation** that higher total steps leads to higher calories burn  
Correlation coefficient of **0.56**  
* The closer to 1.0, the stronger the positive correlation  

```{r user activity scatter plot 2, echo=FALSE}
ggplot(activity_filter_df) + 
  geom_point(aes(total_steps, calories)) +
  geom_smooth(aes(total_steps, calories), 
              method = 'loess', formula = y~x) + 
  labs(title = 'Steps vs Calories', 
       subtitle = 'Correlation between steps and calories', 
       x = 'Total Steps', y = 'Total Calories')
```
```{r}
cor(activity_filter_df$total_steps, activity_filter_df$calories) %>% 
  round(2)
```