---
title: "Google Data Analytics Capstone Project - Bellabeat"
author: "Lin Jiaming"
date: "4 June 2021"
output: html_document
---

## Description

**Activity:** Data Cleaning  
**Dataset:** FitBit Fitness Tracker Data (Kaggle)  
**Table:** dailyActivity_merged.csv

### Load Libraries

```{r library, message=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(skimr)
```

### Import Data

```{r import data, message=FALSE}
activity_df <- read_csv('dailyActivity_merged.csv')
```

### Column Header Check

Manually check column headers using glimpse function  

```{r column header check}
glimpse(activity_df)
```

Clean column headers using clean_names function

```{r column header clean}
activity_df <- clean_names(activity_df)

colnames(activity_df)
```

### Data Type Check

From glimpse, it is observed that:  

* activity_date data type is character  
* id data type is double  

Convert:  

* activity_date data type to date  
* id data type to factor  

```{r data type}
activity_df$activity_date <- mdy(activity_df$activity_date)

activity_df$id <- factor(activity_df$id)
```

### Missing Values Check

No missing values found

```{r missing values}
skim_without_charts(activity_df)
```

### Duplicates Check

No duplicates found

```{r duplicates}
activity_df %>% 
  group_by(id, activity_date) %>% 
  filter(n() > 1)
```

### Extra Space Check

Not applicable  
No character data type

### Inconsistent Field Length Check

Not applicable

### Date Range Check

Start date = 2016-04-12  
End date = 2016-05-12

```{r date range}
activity_df %>% summarise(start_date = min(.$activity_date), 
                          end_date = max(.$activity_date))
```

### Categorization Check

Not applicable

### Data Count Check

There are 940 total records  
There are 33 unique users, and 31 unique dates

```{r data count}
activity_df %>% summarise(record_cnt = nrow(.), 
                          user_cnt = n_distinct(.$id), 
                          date_cnt = n_distinct(.$activity_date), 
                          user_date_cnt = n_distinct(.$id, .$activity_date))
```

Only 21 users tracked their data for all 31 days

```{r users @ 31 days}
activity_df %>% 
  group_by(id) %>% 
  summarise(days_tracked = length(activity_date)) %>% 
  group_by(days_tracked) %>% 
  summarise(user_count = length(id)) %>% 
  arrange(-days_tracked)
```

All users track their data in running days

```{r users @ running days}
# Date_diff > 1 day indicates that users did not track their data in running days
activity_df %>% 
  group_by(id) %>% 
  summarise(date_diff = activity_date - lag(activity_date, 1)) %>% 
  group_by(date_diff) %>% 
  summarise(num = length(id))
```

### Data Range Check

total_steps_min = 0  

total_distance_min = 0  
total_distance_max = 28.03  

```{r data range}
activity_df %>% 
  select(total_steps, total_distance, tracker_distance, logged_activities_distance, calories) %>% 
  summary()
```

There are 77 records where total_steps = 0  
There are 73 records where total_steps = 0, but calories != 0  
**Exclude these records for specific analysis**

```{r data range @ total steps}
activity_df %>% 
  filter(total_steps == 0) %>% 
  nrow()

activity_df %>% 
  filter(total_steps == 0, calories != 0) %>% 
  nrow()
```

There are 78 records where total_distance = 0  
There are 74 records where total_distance = 0, but calories != 0  
**Exclude these records for specific analysis**

```{r data range @ total distance}
activity_df %>% 
  filter(total_distance == 0) %>% 
  nrow()

activity_df %>% 
  filter(total_distance == 0, calories != 0) %>% 
  nrow()
```

There are 77 records where total_distance = 0 and total_steps = 0  
There are 1 records where total_distance = 0 and total_steps != 0

```{r data range @ total steps and total distance}
activity_df %>% 
  filter(total_distance == 0, total_steps == 0) %>% 
  nrow()

activity_df %>% 
  filter(total_distance == 0, total_steps != 0) %>% 
  nrow()
```

Investigate if max of total_distance make sense  
Calculate *speed* to determine if total_distance make sense

Assume units of distance to be in *km*, instead of miles ([Reference](https://dev.fitbit.com/build/reference/web-api/basics/))

Fitbit definition for minutes ([Reference](https://community.fitbit.com/t5/Web-API-Development/Daily-Activity-Summary-Data-Definition-Questions/td-p/3087077)):  

* sedentary_minutes, less than 1.5 METs  
* lightly_active_minutes, between 1.5-3.0 METs  
* fairly_active_minutes, between 3.0-6.0 METs  
* very_active_minutes, greater than 6.0 METs  

Wikipedia definition for MET ([Reference](https://en.wikipedia.org/wiki/Metabolic_equivalent_of_task)):  

* 1.5 METs: writing, deskwork, using computer  
* 2.0 METs: walking slowly

Therefore, to calculate speed, one must *exclude sedentary_minutes*

Speed comparison ([Reference](https://www.runnersworld.com/uk/training/marathon/a27787958/average-marathon-finish-time)):  

* Average marathon runner speed (male) = 60/6.43 = 9.33 km/h  
* Average marathon runner speed (female) = 60/7.26 = 8.26 km/h  
* Speed of fastest human = 37.58 km/h ([Reference](https://www.britannica.com/story/how-fast-is-the-worlds-fastest-human))  

The data's **max speed (very_active) of 10.48 km/h** is **reasonable**  
The data's **max speed (total_distance) of 8.31km/h** is **reasonable**, therefore **total_distance is reasonable** too  

```{r data range @ speed}
activity_df %>% 
  mutate(spd_total = total_distance / (very_active_minutes + 
                                       fairly_active_minutes + 
                                       lightly_active_minutes) * 60) %>% 
  mutate(spd_very_active = very_active_distance / very_active_minutes * 60) %>% 
  mutate(spd_fairly_active = moderately_active_distance / fairly_active_minutes * 60) %>% 
  mutate(spd_lightly_active = light_active_distance / lightly_active_minutes * 60) %>% 
  select(spd_total, spd_very_active, spd_fairly_active, spd_lightly_active) %>% 
  round(2) %>% 
  na_if(Inf) %>% 
  summary()
```

### Cross Field Check

Check total_distance vs tracker_distance  
There are 15 records where total_distance != tracker_distance  
It is unclear what does tracker_distance represents

```{r cross field @ tracker distance}
activity_df %>% 
  filter(total_distance != tracker_distance) %>% 
  mutate(dist_diff = round(total_distance - tracker_distance, 2)) %>% 
  mutate(dist_diff_percent = round(dist_diff / total_distance * 100, 2)) %>% 
  select(id, activity_date, total_distance, tracker_distance, dist_diff, dist_diff_percent) %>% 
  arrange(-dist_diff_percent)
```

Check total_distance vs sum of Distance  
There are 41 records where difference between total_distance and sum of distance > 1%

```{r cross field @ sum of distance}
activity_df %>% 
  mutate(dist_sum = very_active_distance + 
                    moderately_active_distance + 
                    light_active_distance + 
                    sedentary_active_distance) %>% 
  filter(total_distance != dist_sum) %>% 
  mutate(dist_diff = round(total_distance - dist_sum, 2)) %>% 
  mutate(dist_diff_percent = round(dist_diff / total_distance * 100, 2)) %>% 
  select(id, activity_date, total_distance, dist_sum, dist_diff, dist_diff_percent) %>% 
  filter(dist_diff_percent > 1) %>% 
  arrange(-dist_diff_percent, -dist_diff)
```

Check sum of Minutes = 1440 (24 hours * 60 minutes)  
There are 478 records where sum of minutes = 1440  
There are 462 records where sum of minutes != 1440

```{r cross field @ sum of minutes 1}
activity_df %>% 
  mutate(minutes_total = very_active_minutes + 
                         fairly_active_minutes + 
                         lightly_active_minutes + 
                         sedentary_minutes) %>% 
  filter(minutes_total == 1440) %>% 
  #filter(minutes_total != 1440) %>% 
  group_by(minutes_total) %>% 
  summarise(number_of_records = length(id)) %>% 
  summarise(sum_of_records = sum(number_of_records))
```

Check sum of [Minutes in dailyActivity.csv] and [total_time_in_bed in sleepDay.csv] = 1440 (24 hours * 60 minutes)  
There are 604 records where sum of minutes = 1440  
There are 155 records where sum of minutes > 1440  
There are 181 records where sum of minutes < 1440  
Investigation is needed with regards to how Fitbit measure their statistics

```{r cross field @ sum of minutes 2, message=FALSE}
# Import sleepDay.csv
sleep_df <- read_csv('sleepDay_clean.csv')

# Convert id data type to factor
sleep_df$id <- factor(sleep_df$id)

# Manually check column names
glimpse(sleep_df)
```
```{r}
# Sum of [Minutes in dailyActivity.csv] and [TotalTimeInBed in sleepDay.csv]
left_join(activity_df, sleep_df, by = c('id','activity_date' = 'sleep_day')) %>% 
  mutate(minutes_total = very_active_minutes + 
                         fairly_active_minutes + 
                         lightly_active_minutes + 
                         sedentary_minutes + 
                         coalesce(total_time_in_bed,0)) %>% 
  filter(minutes_total == 1440) %>% 
  #filter(minutes_total > 1440) %>% 
  #filter(minutes_total < 1440) %>% 
  group_by(minutes_total) %>% 
  summarise(number_of_records = length(id)) %>% 
  summarise(sum_of_records = sum(number_of_records))
```

### Cross Table Check

All 410 keys (id and date) in sleepDay.csv, are in dailyActivity.csv

```{r cross table}
count(activity_df)
count(sleep_df)
inner_join(activity_df, sleep_df, by = c('id','activity_date' = 'sleep_day'))
```

### Export Data

Export table with all records and fields  
Data where total_distance = 0 and total_steps = 0 will be filtered for specific analysis

```{r export data}
write_csv(activity_df, 'dailyActivity_clean.csv')
```