---
title: "Google Data Analytics Capstone Project - Bellabeat"
author: "Lin Jiaming"
date: "4 June 2021"
output: html_document
---

## Description

**Activity:** Data Cleaning  
**Dataset:** FitBit Fitness Tracker Data (Kaggle)  
**Table:** sleepDay_merged.csv

### Load Libraries

```{r library, message=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(skimr)
```

### Import Data

```{r import data, message=FALSE}
sleep_df <- read_csv('sleepDay_merged.csv')
```

### Column Header Check

Manually check column headers using glimpse function  

```{r column header check}
glimpse(sleep_df)
```

Clean column headers using clean_names function

```{r column header clean}
sleep_df <- clean_names(sleep_df)

colnames(sleep_df)
```

### Data Type Check

From glimpse, it is observed that:  

* sleep_day data type is character  
* id data type is double  
* total_sleep_records data type is double  

Convert:  

* sleep_day data type to date  
* id data type to factor  
* total_sleep_records data type to factor  

```{r data type}
sleep_df$sleep_day <- sleep_df$sleep_day %>% 
  mdy_hms() %>% 
  as_date()

sleep_df$id <- factor(sleep_df$id)
sleep_df$total_sleep_records <- factor(sleep_df$total_sleep_records)
```

### Missing Values Check

No missing values found

```{r missing values}
skim_without_charts(sleep_df)
```

### Duplicates Check

Removed 3 duplicates

```{r duplicates}
# Duplicates are exactly the same in all fields
sleep_df %>% 
  group_by(id, sleep_day) %>% 
  filter(n() > 1)
```
```{r duplicates removal}
sleep_df <- distinct(sleep_df, id, sleep_day, .keep_all = TRUE)
```

### Extra Space Check

Not applicable  
No character data type

### Inconsistent Field Length Check

Not applicable

### Date Range Check

Start date = 2016-04-12  
End date = 2016-05-12

```{r date range}
sleep_df %>% summarise(start_date = min(.$sleep_day), 
                       end_date = max(.$sleep_day))
```

### Categorization Check

3 categories for total_sleep_records

```{r categorization}
distinct(sleep_df, total_sleep_records)
```

### Data Count Check

There are 410 total records  
There are 24 unique users, and 31 unique dates  
Not all users participated in tracking sleep

```{r data count}
sleep_df %>% summarise(record_cnt = nrow(.), 
                       user_cnt = n_distinct(.$id), 
                       date_cnt = n_distinct(.$sleep_day), 
                       user_date_cnt = n_distinct(.$id, .$sleep_day))
```

Only 3 users tracked their data for all 31 days

```{r users @ 31 days}
sleep_df %>% 
  group_by(id) %>% 
  summarise(days_tracked = length(sleep_day)) %>% 
  group_by(days_tracked) %>% 
  summarise(user_count = length(id)) %>% 
  arrange(-days_tracked)
```

19 out of 24 users did not track their data in running days  
For older fitbits and most "pocket" models, you have to press a button ([Reference](https://www.quora.com/How-does-Fitbit-figure-out-that-Im-sleeping))

```{r users @ running days}
# Date_diff > 1 day indicates that users did not track their data in running days
sleep_df %>% 
  group_by(id) %>% 
  summarise(date_diff = sleep_day - lag(sleep_day, 1)) %>% 
  filter(date_diff != 1) %>% 
  distinct(id) %>% 
  nrow()
```

### Data Range Check

total_minutes_asleep_max = 796min = 13.3 hours  
total_minutes_asleep_min = 58min = 0.97 hours  

total_time_in_bed_max = 961mins = 16 hours  
total_time_in_bed_min = 61mins = 1.02 hours

```{r data range}
sleep_df %>% 
  select(total_minutes_asleep, total_time_in_bed) %>% 
  summary()
```

### Export Data

Export table with all records and fields

```{r export data}
write_csv(sleep_df, 'sleepDay_clean.csv')
```